Novel genetic control of migratory diapause in Australian Monarch Butterflies

### 2024-01-04 paper/paper_draft.Rmd
updated paper with new association approach

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(snpR); library(data.table); library(ggplot2); library(gridExtra); library(dplyr); library(openxlsx); library(jpeg); library(nlme)

@@

```{r import_and_diversity, echo=FALSE, message=FALSE, warning=FALSE,include=FALSE}
L <- 6578062 # total number of sequenced bases

#==========prep data==========
# read in genotypes and metadata
sample.meta <- read.table("../data/sample_metadata.txt", header = T, stringsAsFactors = F)
genos <- read.table("../data/v4/genotypes.geno", stringsAsFactors = F)
snp.meta <- genos[,1:2]
genos <- genos[,-c(1:2)]
colnames(snp.meta) <- c("scaffold", "position")

# rename msc scaffolds
chrs <- which(grepl("chr",snp.meta$scaffold))
snp.meta$scaffold[-chrs] <- "other"
snp.meta$scaffold <- factor(snp.meta$scaffold, levels = c(paste0("chr", 1:30), "other"))

# remove samples with no phenotypic data
nas <- which(is.na(sample.meta$likely_pheno))
genos <- genos[,-which(is.na(sample.meta$likely_pheno))]
sample.meta <- sample.meta[-which(is.na(sample.meta$relaxed_pheno)),]

# add in colony info
fam_info <- sample.meta(readRDS("../colony/colony_R_outputs.RDS")$col_res$x)
fam_info$MotherID <- as.numeric(gsub("#", "", fam_info$MotherID))
fam_info$MotherID <- factor(fam_info$MotherID, levels = 1:max(fam_info$MotherID))
sample.meta$MotherID <- fam_info$MotherID # note, individuals are in identical order.

# send year to factor, import, and filter
sample.meta$year <- as.factor(sample.meta$year)
# dat <- import.snpR.data(genos, snp.meta, sample.meta)
# # dat <- dat[,-which(sample.meta(dat)$year != 2018)] # note: same locus pops out if you only use my samples, but is stronger with both. Good sign!
# dat <- filter_snps(dat, hwe = 0.000001, hwe_facets = "year")

#==========diversity===========
# dat <- calc_pi(dat, c(".base", "year"))
# dat <- calc_ho(dat, c(".base", "year"))
# dat <- calc_fis(dat, c(".base", "year"))
# dat <- calc_pairwise_fst(dat, "year", boot = 100, boot_par = 6, verbose = TRUE)
# saveRDS(dat, "diversity_stats.RDS")
dat <- readRDS("diversity_stats.RDS")
diversity <- get.snpR.stats(dat, c(".base", "year"), c("pi", "ho", "fis"))$weighted.means
diversity$weighted_mean_pi <- diversity$weighted_mean_pi/L
diversity$weighted_mean_ho <- diversity$weighted_mean_ho/L
fst <- get.snpR.stats(dat, "year", "fst")$weighted.means

#==========PCA===============
pca <- plot_clusters(dat, c("year.MotherID")) # color by mother and by year

@@

```{r association, echo=FALSE, warning=FALSE, message=FALSE}
# family bias
bias <- sample.meta[,c("MotherID", "likely_pheno")]
bias$MotherID <- as.factor(bias$MotherID)
bias$likely_pheno <- as.factor(bias$likely_pheno)
fisher_res <- fisher.test(bias$MotherID, bias$likely_pheno, simulate.p.value = TRUE)

bias_p <- reshape2::dcast(bias, MotherID ~ likely_pheno)
bias_p$total <- bias_p$M + bias_p$R
bias_p$proportion <- bias_p$M/bias_p$total
# Figure3a <- ggplot(bias, aes(x = bias$MotherID, fill = bias$likely_pheno)) + geom_bar(position = 'dodge')
Figure3a <- ggplot(bias_p, aes(x = MotherID, y = proportion, size = total)) + geom_point() + theme_bw() + geom_hline(yintercept = .5) + xlab("Maternal Family") + ylab("Proportion Diapaused") + labs(size = "n")
Figure3al <- ggpubr::get_legend(Figure3a)
Figure3a <- Figure3a + theme(legend.position = "none")

# association test
dat <- calc_association(dat, response = "likely_pheno", maxiter = 1000)

# top SNP
top_snp <- dplyr::arrange(get.snpR.stats(dat, stats = c("association", "maf"))$single, gmmat_pval_likely_pheno)[1,]
match <- which(snp.meta(dat)$scaffold == top_snp$scaffold[1] & snp.meta(dat)$position == top_snp$position[1])
gmatch <- genotypes(dat)[match,]
gmatch <- data.frame(genotype = unlist(gmatch), phenotype = sample.meta(dat)$likely_pheno)
gmatch$genotype[gmatch$genotype == "NN"] <- NA
gmatch$genotype <- ifelse(gmatch$genotype == "GG", "resident", "migrant")
gtable <- table(gmatch)
gtable <- addmargins(gtable)

# plot
Figure3b <- plot_manhattan(dat, "gmmat_pval_likely_pheno", chr = "scaffold", significant = 0.00005, suggestive = 0.0005, log.p = T,
                           chr_order = c(paste0("chr", 1:30), "other"), highlight_style = "red") # highlight top snp
Figure3b <- Figure3b$plot + ylab("-log10(p-value)") + xlab("Position") + theme(strip.text = element_blank())
```

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(snpR); library(data.table); library(ggplot2); library(dplyr);
library(cowplot); library(openxlsx); library(nlme); library(gggenes);
library(Biostrings)

@@

```{r import_and_diversity, echo=FALSE, message=FALSE, warning=FALSE,include=FALSE}
#==========prep data==========
# read in genotypes and metadata
sample.meta <- read.table("../data/sample_metadata_v4.txt", header = T, stringsAsFactors = F)

# remove samples with no phenotypic data
# nas <- which(is.na(sample.meta$likely_pheno))
# genos <- genos[,-which(is.na(sample.meta$likely_pheno))]
# sample.meta <- sample.meta[-which(is.na(sample.meta$relaxed_pheno)),]

# add in colony info
fam_info <- sample.meta(readRDS("../colony/colony_R_outputs.RDS")$col_res$x)
fam_info$MotherID <- as.numeric(gsub("#", "", fam_info$MotherID))
fam_info$MotherID <- factor(fam_info$MotherID, levels = 1:max(fam_info$MotherID))
sample.meta$MotherID <- fam_info$MotherID[match(sample.meta$SampleID, fam_info$SampleID)] 
fam_info$FatherID <- as.numeric(gsub("\\*", "", fam_info$FatherID))
fam_info$FatherID <- factor(fam_info$FatherID, levels = 1:max(fam_info$FatherID))
sample.meta$FatherID <- fam_info$FatherID[match(sample.meta$SampleID, fam_info$SampleID)] 
sample.meta$cluster <- fam_info$ClusterIndex[match(sample.meta$SampleID, fam_info$SampleID)]

# send year to factor, import, and filter
sample.meta$year <- as.factor(sample.meta$year)

#==========PCA plot==============
PCA <- fread("../scripts/association/PCAngsd_out.cov")
e <- eigen(as.matrix(PCA))
PCA <- e$vectors
PCA <- as.data.table(PCA)
colnames(PCA) <- paste0("PC", 1:ncol(PCA))
t.sizes <- c(18, 12)
FigureS1 <- ggplot(PCA, aes(x = PC1, y = PC2)) +
  geom_point(size = 3) +
  theme_bw() + 
  theme(legend.text = element_text(size = t.sizes[2]),
        axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        axis.text.y = element_text(size = t.sizes[2]),
        legend.title = element_text(size = t.sizes[1])) +
  xlab(paste0("PC1 (", round(e$values[1]/sum(e$values)*100, 3), "%)")) +
  ylab(paste0("PC2 (", round(e$values[2]/sum(e$values)*100, 3), "%)"))

@@

```{r Figure2, echo=FALSE, warning=FALSE, message=FALSE}
#==========family bias===================
bias <- sample.meta[,c("MotherID", "FatherID", "quant_pheno")]
bias$quant_pheno <- bias$quant_pheno/2
bias$MotherID <- as.factor(bias$MotherID)
bias$FatherID <- as.factor(bias$FatherID)

# permute to show expected random range per rank
perms <- matrix(0, 1000, length(unique(bias$MotherID)))
for(i in 1:nrow(perms)){
  tb <- bias
  tb$quant_pheno <- sample(tb$quant_pheno, nrow(tb), TRUE)
  perm_p <- tapply(tb$quant_pheno, tb$MotherID, mean)
  perm_p <- as.data.frame(perm_p)
  perm_p$MotherID <- 1:nrow(perm_p)
  colnames(perm_p)[1] <- "score"
  perm_p$rank <- rank(perm_p$score, ties.method = "first")
  perm_p$total <- as.numeric(table(tb$MotherID))
  perms[i,] <- perm_p$score[order(perm_p$rank)]
}
perm_summary <- data.table(score = colMeans(perms),
                           rank = 1:ncol(perms))
perm_summary[, c("lCI", "uCI") := as.data.frame(t(apply(perms, MARGIN = 2, FUN = function(x){
  return(t.test(x)$conf.int)
})))]

# anova to show significance
m1 <- lm(quant_pheno ~ MotherID, bias)
m1_an <- anova(m1)

# convert for plotting
bias_p <- tapply(bias$quant_pheno, bias$MotherID, mean)
bias_p <- as.data.frame(bias_p)
bias_p$MotherID <- 1:nrow(bias_p)
colnames(bias_p)[1] <- "score"
bias_p$rank <- rank(bias_p$score, ties.method = "first")
bias_p$total <- as.numeric(table(bias$MotherID))

# generate plot
t.sizes <- c(18, 12)
Figure2a <- ggplot(bias_p, aes(x = rank, y = score, size = total)) +
  geom_point(color ="#393e42") + theme_bw() +
  geom_line(data = perm_summary, aes(x = rank, y = score), inherit.aes = FALSE, color = "#eb612e") +
  geom_ribbon(data = perm_summary, aes(ymin = lCI, ymax = uCI, x = rank), 
              inherit.aes = FALSE, color = "#eb612e", alpha = .2) +
  xlab("Maternal Family") +
  ylab("Mean Reproductive Score") +
  labs(size = "Family Size")  +
  theme(legend.text = element_text(size = t.sizes[2]),
        axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        axis.text.y = element_text(size = t.sizes[2]),
        legend.title = element_text(size = t.sizes[1])) +
  scale_size_continuous(range = c(5, 10))

# ggsave("~/2022-2023/ICCB2023/scatter.jpg", Figure3a, device = "jpg", dpi = 1000, height = 4, width = 8, scale = 3)

#==========association test -- read in from ANGSD============
chr_key <- fread("../data/chr_info.tsv")
chr_key$Chr_ID <- chr_key$`Chromosome name`
chr_key$Chr_ID <- make.unique(chr_key$Chr_ID)
chr_key$partID <- as.numeric(gsub(".+\\.", "", chr_key$Chr_ID))
chr_key$partID[-grep("\\.", chr_key$Chr_ID)] <- 0
chr_key <- dplyr::arrange(chr_key, as.numeric(`Chromosome name`), partID)
adat_quant <- data.table::fread("../scripts/association/association_out_quant_cov.lrt0.gz", header = T)
adat_quant$chr <- chr_key$Chr_ID[match(adat_quant$Chromosome, chr_key$`RefSeq seq accession`)]
adat_quant[grep("Un", chr), chr := "Other"]
adat_quant$pval <- pchisq(adat_quant$LRT, 1, lower.tail = FALSE)
adat_quant$lpval <- -log(adat_quant$pval)
adat_quant$fdr <- p.adjust(adat_quant$pval, "fdr")
adat_quant <- dplyr::arrange(adat_quant, pval)

q_quant <- plot_qq(adat_quant, "pval")
Figure2b <- ggplot(q_quant$.base$data, aes(x = .e, y = .o)) +
  geom_point(color = "#393e42") +
  geom_abline(slope = 1, intercept = 0, color = "#eb612e") +
  theme_bw() +
  theme(legend.text = element_text(size = t.sizes[2]),
        axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        axis.text.y = element_text(size = t.sizes[2]),
        legend.title = element_text(size = t.sizes[1])) +
  ylab(bquote(Observed ~ log[10](p))) +
  xlab(bquote(Expected ~ log[10](p)))

# note -- color scheme for posters!
ord <- unique(chr_key$Chr_ID, sort = F)
ord <- ord[-grep("Un", ord)]
ord <- ord[-which(ord == "Z")]
ord <- c("Z", ord, "Other")
Figure2c <- plot_manhattan(adat_quant, "fdr", chr = "chr", bp = "Position",
                           significant = 0.05, suggestive = .1, log.p = TRUE,
                           chr_order = ord,
                           highlight = FALSE, colors = c("#eb612e", "#393e42"), 
                           t.sizes = c(t.sizes[1], t.sizes[1], t.sizes[2]), 
                           simplify_output = TRUE) # highlight top snp

# refetch chr centers, pulled from snpR
chr.info <- tapply(Figure2c$data[,"cum.bp"], Figure2c$data[,"chr"], max) # chromosome lengths
chr.centers <- rowMeans(cbind(tapply(Figure2c$data[,"cum.bp"], Figure2c$data[,"chr"], min), chr.info)) # chromosome centers

# plot
Figure2c <- Figure2c +
  ylab("-log10(p-value)") +
  xlab("Chromosome") +
  theme(legend.text = element_text(size = t.sizes[2]),
        axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        axis.text.y = element_text(size = t.sizes[2]),
        legend.title = element_text(size = t.sizes[1])) +
  scale_x_continuous(expand = c(0, 0), breaks = chr.centers, labels = 
                       names(chr.centers))

# ggsave("~/2022-2023/ICCB2023/manhattan.jpg", Figure3b, device = "jpg", dpi = 1000, height = 4, width = 8, scale = 3)

#==========combine=======================
Figure2_top <- plot_grid(Figure2a, Figure2b, align = "hv", axis = "tl")

Figure2 <- plot_grid(Figure2_top,
                     Figure2c,
                     nrow = 2)

save_plot("Figure2.pdf", Figure2, base_height = 11, base_width = 15)

@@

```{r genes_of_interest}
#=======identify targets=========
window <- 50*1000 # 10kb
targets <- adat_quant[fdr <= 0.05,]
targets[,start := Position - window]
targets[,end := Position + window]
targets <- dplyr::arrange(targets, Chromosome, start)

# one peak per chr, so simplify
targets[,start := min(start), by = chr]
targets[,end := max(end), by = chr]
targets <- unique(targets[,c("Chromosome", "chr", "start", "end")])

#=======locate genes=============
gtf <- rtracklayer::readGFF("../data/genomic.gtf")
gtf <- as.data.table(gtf)
gtf[,direction := ifelse(strand == "+", TRUE, FALSE)]
keep <- vector("list", nrow(targets))
all_genes <- keep
for(i in 1:nrow(targets)){
  keep[[i]] <- gtf[which(seqid == targets[i,Chromosome] &
                     ((end >= targets[i,start] & end <= targets[i,end]) |
                        (start <= targets[i,end] & start >= targets[i,start]) |
                        (start <= targets[i, start] & end >= targets[i, end]))),]
  all_genes[[i]] <- unique(keep[[i]][type == "gene",])
}
all_genes <- rbindlist(all_genes, idcol = "target")
IDs <- gsub("GeneID:", "", all_genes$db_xref)

# get the prot sequences from NCBI
cmd <- paste0("~/bin/datasets download gene gene-id ", paste0(IDs, collapse = " "))
system(cmd)

# copy the .faa and paste into monarchbase, read in the result here
prot_matches <- fread("../data/prot_matches_monarchbase.txt")
prot_matches[,LOC := unlist(stringi::stri_extract_all_regex(`Query ID`, "LOC.{9}"))]
prot_matches <- dplyr::arrange(prot_matches, LOC, Evalue)
prot_matches <- prot_matches[, head(.SD, 1), by= `Query ID`]
prot_matches <- dplyr::arrange(prot_matches, Length)
prot_matches <- prot_matches[, head(.SD, 1), by= LOC]

all_genes <- merge(all_genes, prot_matches, by.x = "gene_id", by.y = "LOC", all = TRUE)

#========plot====================
plot_parts <- vector("list", nrow(targets))
for(i in 1:nrow(targets)){
  # figure out max and min range
  min_pos <- targets[i,start]
  max_pos <- targets[i,end]

  tgtf <- keep[[i]]
  subgene <- tgtf[type == "CDS",]
  tgtf <- tgtf[type == "gene",]
  subgene[,gene_start := tgtf$start[match(gene, tgtf$gene)]]
  subgene[,gene_end := tgtf$end[match(gene, tgtf$gene)]]

  tgtf[,gene_lab := start + ((end - start)/2)]
  tgtf$OGS_name <- all_genes$`Monarch geneid`[match(tgtf$gene_id, all_genes$gene_id)]
  tgtf$OGS_name[is.na(tgtf$OGS_name)] <- tgtf$gene_id[is.na(tgtf$OGS_name)]

  subgene[,OGS_name := tgtf$OGS_name[match(gene, tgtf$gene)]]

  if(min_pos > min(c(tgtf$start, tgtf$end))){
    min_pos <- min(c(tgtf$start, tgtf$end))
  }
  if(max_pos < max(c(tgtf$start, tgtf$end))){
    max_pos <- max(c(tgtf$start, tgtf$end))
  }

  # plot SNPs

  tsnps <- adat_quant[which(Chromosome == targets[i,Chromosome] &
                     Position >= min_pos & Position <= max_pos),]

  pt <- ggplot(tsnps, aes(x = Position/1000000, y = -log10(fdr))) +
    geom_point() +
    theme_bw() +
    scale_x_continuous(limits = c(min_pos/1000000, max_pos/1000000)) +
    scale_y_continuous(expand = c(0, .05))

  if(i == 1)
    pt <- pt + 
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.text.y = element_text(size = t.sizes[2]),
          axis.title.y = element_text(size = t.sizes[1]),
          axis.ticks.x = element_blank()) +
  ylab(bquote(-log[10](p[FDR])))

  if(i != 1){
    pt <- pt + 
      theme(axis.text.x = element_blank(),
          axis.title = element_blank(),
          axis.text.y = element_text(size = t.sizes[2]),
          axis.ticks.x = element_blank())
  }

  pb <- ggplot(tgtf, aes(xmin = start/1000000, xmax = end/1000000, 
                         y = OGS_name, forward = direction, fill = gene)) +
    geom_gene_arrow(arrow_body_height = unit(1, "lines"),
                    arrowhead_height = unit(1.5, "lines"), show.legend = FALSE,
                    fill = "#E5C16F") + 
    geom_subgene_arrow(data = subgene, 
                       aes(xmin = gene_start/1000000, xmax = gene_end/1000000, 
                           xsubmin = start/1000000, xsubmax = end/1000000,
                           y = OGS_name), 
                       fill = "black", arrow_body_height = unit(1, "lines"),
                       arrowhead_height = unit(1.5, "lines")) +
    theme_bw() +
    scale_x_continuous(limits = c(min_pos/1000000, max_pos/1000000)) +
    ylab("Gene") +
    xlab(paste0("Chromosome ", targets[i]$chr, "Position (kb)"))

  if(i == 1)
    pb <- pb + 
    theme(axis.text = element_text(size = t.sizes[2]),
          axis.title = element_text(size = t.sizes[1]))

  if(i != 1){
    pb <- pb + 
      theme(axis.title.y = element_blank(),
          axis.text = element_text(size = t.sizes[2]),
          axis.title.x = element_text(size = t.sizes[1]))
  }

  plot_parts[[i]] <- plot_grid(pt, pb, ncol = 1, rel_heights = c(1, 1),
                               align = "hv", axis = "tl")

}

Figure3 <- plot_grid(plot_parts[[1]],
                     plot_parts[[2]] + theme(axis.title.y = element_blank()),
                     plot_parts[[3]] + theme(axis.title.y = element_blank()), nrow = 1) +
  theme(plot.margin = unit(c(.1, .5, .1, .5), "cm"))

save_plot("Figure3.pdf", Figure3, base_height = 11, base_width = 15)
